### 2015-06-17 chapters/ipa_meets_unicode.tex
update section diacritic ordering
I removed most references to PHOIBLE here. That is more a use case for
later. This section should be about “strict” IPA

A similar problem arises with the rhotic hook. There are two precomposed
characters in Unicode with a rhotic hook, which are not canonical equivalent 
with a combination of the vowel with a separately encoded hook:
\begin{itemize}
  \item[] <ɚ> \textsc{latin small letter schwa with hook} at \uni{025A}
  \item[] <ɝ> \textsc{latin small letter reversed open e with hook}at \uni{025D}
\end{itemize}

@@

Another pitfall relates the concept of what are diacritics. The problem is that
the meaning of the term diacritics as used by the IPA is not the same as is used
in the Unicode Standard. Specifically, diacritics in the IPA-sense are either
so-called \textsc{combining diacritical marks} or \textsc{spacing modifier
letters} in the Unicode Standard. Crucially, Combining Diacritical Marks are by
definition combined with the character before them (to form so-called default
grapheme clusters, see Section~\ref{the-unicode-approach}). In contrast, Spacing
Modifier Letters are by definition \emph{not} combined into grapheme clusters
with the preceding character, but simply treated as separate letters. In the
context of the IPA, the following IPA-diacritics are actually Spacing Modifier
Letters in the Unicode Standard:

@@

	\item[] Superscript letters, like:
	\begin{itemize}
	  \item[] <ʰ> \textsc{modifier letter small h} at \uni{02B0}
	  \item[] <ˤ> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}
	  \item[] and many more like this
	\end{itemize}

@@

Facing the problem of specifying a consistent ordering of diacritics while
developing a large database of phonological inventories from the world's
languages, \citet[540]{Moran2012} defines a set of diacritic ordering
conventions. As new languages' phonological inventories are added to the 
database,\footnote{\url{http://phoible.org/}} updates and additions to the 
conventions are published online by~\cite{MoranMcCloy2014}.\footnote{
\url{http://phoible.github.io/conventions/}} The conventions are influenced by
the linguistic literature, though some ad-hoc decisions had to be taken, like 
introducing several symbols for widely attested sounds that lack representation 
in the current version of the IPA.\@ Whenever possible, additions were drawn from 
the extIPA symbols for disordered speech.\footnote{\url{https://www.internationalphoneticassociation.org/sites/default/files/extIPAChart2008.pdf}}

\cite{MoranMcCloy2014} explicitly define all character sequences so that the vast variety
of phonemes found in a large sample of descriptions of the world's language were normalized into
consistent character sequences, e.g.~if one language description uses <kʲʼ> and
another <kʼʲ>, when both are intended to be phonetically equivalent, then a decision
to normalize to one form was taken. For example, when a character sequence
contains more than one character in Spacing Modifier Letters, the order that is
proposed is the following (where <\textbar{}> indicates \textit{or} and <→>
indicates \textit{precedes}):

\begin{itemize}
	\item[] \textsc{Spacing Modifier Letters ordering:}
	\begin{itemize}
	  \item[→] (unreleased <\ \ ̚ >)
	  \item[→] (rhotic hook <\ \ ˞ >)
	  \item[→] (lateral release <ˡ> \textbar{} nasal release <ⁿ>)
	  \item[→] (labialized <ʷ>)
	  \item[→] (palatalized <ʲ>)
	  \item[→] (velarized <ˠ>)
	  \item[→] (pharyngealized <ˤ>)
	  \item[→] (glottalized <ˀ>)
	  \item[→] (aspirated <ʰ> \textbar{} ejective <ʼ>)
	  \item[→] (long <ː> \textbar{} half-long <ˑ>)
	\end{itemize}
\end{itemize}

@@

\begin{itemize}
	\item[] \textsc{Combining Diacritical Marks (below) ordering:}
	\begin{itemize}	
	  \item[→] (linguolabial <t̼> \textbar{} dental <t̪> \textbar{} apical <t̺> \textbar{} laminal <t̻>)
	  \item[→] (advanced <u̟> \textbar{} retracted <e̠>)	 
	  \item[→] (raised <e̝> \textbar{} lowered <e̞>) 
	  \item[→] (advanced tongue root <e̘> \textbar{} retracted tongue root <e̙>)
	  \item[→] (frictionalized <ɾ͓>)
	  \item[→] (more round <ɔ̹> \textbar{} less round <ɔ̜>)
	  \item[→] (creaky voice <b̰> \textbar{} breathy voice <b̤> \textbar{} voiced <s̬> \textbar{} devoiced <n̥>)
	  \item[→] (syllabic <n̩> \textbar{} non-syllabic <e̯>)
	\end{itemize}
 \end{itemize}

Character sequences with diacritics above the base character were not
problematic in \citet{Moran2012} because they include only the centralized,
mid-centralized and nasalized combining characters.

\citet{Moran2012,MoranMcCloy2014} mark
tones as singletons with Space Modifier Letters, e.g. \textless{}˦\textgreater{}
for a phonemic high tone, instead of accent diacritics, alleviating potential
conflicts. Building on these works, if a character sequence
contains more than one diacritic above the base character, we propose:

\begin{itemize}
	\item[] \textsc{Combining Diacritical Marks (above) ordering:}
	\begin{itemize}
	  \item[→] (centralized <ë> \textbar{} mid-centralized <e̽>)
	  \item[→] (extra short <ĕ>)
	  \item[→] (tone accents, e.g. <è>)
	  \item[→] (Spacing Modifier Letters, see above)
	  \item[→] (tone letters, e.g. <˦>)
	\end{itemize}	 
\end{itemize}

>>>>

A similar problem arises with the rhotic hook. There are two precomposed
characters in Unicode with a rhotic hook, which are not canonical equivalent 
with a combination of the vowel with a separately encoded hook:
\begin{itemize}
  \item[] <ɚ> \textsc{latin small letter schwa with hook} at \uni{025A}
  \item[] <ɝ> \textsc{latin small letter reversed open e with hook} at \uni{025D}
\end{itemize}

@@

Another pitfall relates to the question of what are diacritics. The problem is that
the meaning of the term diacritics as used by the IPA is not the same as is used
in the Unicode Standard. Specifically, diacritics in the IPA-sense are either
so-called \textsc{combining diacritical marks} or \textsc{spacing modifier
letters} in the Unicode Standard. Crucially, Combining Diacritical Marks are by
definition combined with the character before them (to form so-called default
grapheme clusters, see Section~\ref{the-unicode-approach}). In contrast, Spacing
Modifier Letters are by definition \emph{not} combined into grapheme clusters
with the preceding character, but simply treated as separate letters. In the
context of the IPA, the following IPA-diacritics are actually Spacing Modifier
Letters in the Unicode Standard:

@@

	\item[] Superscript letters, like:
	\begin{itemize}
	  \item[] <ʰ> \textsc{modifier letter small h} at \uni{02B0}
	  \item[] <ˤ> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}
	  \item[] <ⁿ> \textsc{superscript latin small letter n} at \uni{207F}
	  \item[] and many more like this
	\end{itemize}

@@

Facing the problem of specifying a consistent ordering of diacritics while
developing a large database of phonological inventories from the world's
languages, \citet[540]{Moran2012} defines a set of diacritic ordering
conventions. The conventions are influenced by the linguistic literature, though
some ad-hoc decisions had to be taken. For example, when a character sequence
contains more than one character in Spacing Modifier Letters, the order that is
proposed is the following (where <\textbar{}> indicates \textit{or} and <→>
indicates \textit{precedes}):

\begin{itemize}
	\item[] \textsc{Spacing Modifier Letters ordering:}
	\begin{itemize}
	  \item[→] rhotic hook <{\large \ ˞} >
	  \item[→] lateral release <ˡ> \textbar{} nasal release <ⁿ>
	  \item[→] labialized <ʷ>
	  \item[→] palatalized <ʲ>
	  \item[→] velarized <ˠ>
	  \item[→] pharyngealized <ˤ>
	  \item[→] glottalized <ˀ>
	  \item[→] aspirated <ʰ> \textbar{} ejective <ʼ>
	  \item[→] long <ː> \textbar{} half-long <ˑ>
	\end{itemize}
\end{itemize}

@@

\begin{itemize}
	\item[] \textsc{Combining Diacritical Marks (below) ordering:}
	\begin{itemize}	
	  \item[→] linguolabial <t̼> \textbar{} dental <t̪> \textbar{} apical <t̺> \textbar{} laminal <t̻>
	  \item[→] advanced <u̟> \textbar{} retracted <e̠> 
	  \item[→] raised <e̝> \textbar{} lowered <e̞>
	  \item[→] advanced tongue root <e̘> \textbar{} retracted tongue root <e̙>
	  \item[→] frictionalized <ɾ͓>
	  \item[→] more round <ɔ̹> \textbar{} less round <ɔ̜>
	  \item[→] creaky voice <b̰> \textbar{} breathy voice <b̤> \textbar{} voiced <s̬> \textbar{} devoiced <n̥>
	  \item[→] syllabic <n̩> \textbar{} non-syllabic <e̯>
	\end{itemize}
 \end{itemize}

Finally, if a character sequence contains more than one diacritic above the base
character, we propose the following order, with Spacing Modifiers and tone
letters being added at the end:

\begin{itemize}
	\item[] \textsc{Combining Diacritical Marks (above) ordering:}
	\begin{itemize}
	  \item[→] centralized <ë> \textbar{} mid-centralized <e̽>
	  \item[→] extra short <ĕ>
%	  \item[→] (tone accents, e.g. <è>)
	  \item[→] unreleased <e̚>
	  \item[→] Spacing Modifier Letters, see above
	  \item[→] tone letters <˥ ˦ ˧ ˨ ˩>
	\end{itemize}	 
\end{itemize}

### 2015-06-17 chapters/use_cases.tex
reverting this morning's width adjustments
update section diacritic ordering
I removed most references to PHOIBLE here. That is more a use case for
later. This section should be about “strict” IPA

>>>>

\section{PHOIBLE}

As new languages' phonological inventories are added to the 
database,\footnote{\url{http://phoible.org/}} updates and additions to the 
conventions are published online by~\cite{MoranMcCloy2014}.\footnote{
\url{http://phoible.github.io/conventions/}} , like 
introducing several symbols for widely attested sounds that lack representation 
in the current version of the IPA.\@ Whenever possible, additions were drawn from 
the extIPA symbols for disordered speech.\footnote{\url{https://www.internationalphoneticassociation.org/sites/default/files/extIPAChart2008.pdf}}

\cite{MoranMcCloy2014} explicitly define all character sequences so that the
vast variety of phonemes found in a large sample of descriptions of the world's
language were normalized into consistent character sequences, e.g.~if one
language description uses <kʲʼ> and another <kʼʲ>, when both are intended to be
phonetically equivalent, then a decision to normalize to one form was taken.

Character sequences with diacritics above the base character were not
problematic in \citet{Moran2012} because they include only the centralized,
mid-centralized and nasalized combining characters.

\citet{Moran2012;MoranMcCloy2014} mark tones as singletons with Space Modifier
Letters, e.g. <˦> for a phonemic high tone, instead of accent diacritics,
alleviating potential conflicts. Building on these works,
